\documentclass[]{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{listings}
\usepackage{multirow}
\usepackage{tikz}
\usepackage{tikz-qtree}
\usepackage{tipa}
\usetikzlibrary{arrows,automata}
\begin{document}
\newcommand*{\xml}[1]{\texttt{<#1>}}
\theoremstyle{definition}
\newtheorem{thm}{Theorem}
\title{COMS W3261 \\ Computer Science Theory \\ Lecture 16 \\ Post's
Correspondence Problem}
\author{Alexander Roth}
\date{2014 -- 10 -- 29}
\maketitle

\section*{Outline}
\begin{enumerate}
\item Review
\item Post's correspondence problem
\item Modified PCP
\item Undecidability of the ambiguity problem for CFG's
\end{enumerate}

\section{Review}
\begin{itemize}
\item The diagonal language $L_d$ is not recursively enumerable.
\item The universal language $L_u$ is recursively enumerable but not recursive.
\item The complement of the diagonal language is recursively enumerable but not
recursive.
\item The complement of the universal language is not recursively enumerable.
\end{itemize}

\section{Post's Correspondence Problem (PCP)}
\begin{itemize}
\item An instance of Post's correspondence problem consists of two lists of
strings over some alphabet where the two lists have the same number of strings.
Let $A = (w_1, w_2, \ldots, w_k)$ and $B = (x_1, x_2, \ldots, x_k)$ be the two
lists.
\item A solution to this instance of PCP, if one exists, is a sequence of one or
more integers $i_1, i_2, \ldots, i_m$ such that $w_{i_1} w_{i_2} \ldots w_{i_m}
= x_{i_1} x_{i_2}\ldots x_{i_m}$.
\item Example: Let $A = (a, b, ca, abc)$ and $B = (ab, ca, a, c)$. The sequence
$1, 2, 3, 1, 4$ is a solution because the same string $abcaaabc$ is obtained by
concatenating the corresponding strings from either list $A$ $\lbrack
(a)(b)(ca)(a)(abc)\rbrack$.
\item Post's correspondence problem is to determine whether an instance has a
solution.
\item We will show that Post's correspondence problem is undecidable by reducing
the universal language to PCP.
\item We will then show that the ambiguity problem for CFG's is undecidable by
reducing PCP to the ambiguity problem for CFG's.
\end{itemize}

\section{Modified PCP}
\begin{itemize}
\item The Modified PCP has the additional requirement that the first string from
list $A$ and the first string from list $B$ has to be the first string in the
solution. The example above has this property.
\item Formally, a solution to an instance of the MPCP is a sequence of zero or
more integers $i_1, i_2, \ldots, i_m$ such that
$w_1w_{i_1}w_{i_2}\ldots{w_{i_m}} = x_1x_{i_1}x_{i_2}\ldots{x_{i_m}}$.
\item We can show that the PCP problem can be reduced to the modified PCP
problem as follows:
\begin{itemize}
\item We are given an instance of MPCP with lists $A = (w_1,w_2, \ldots, w_k)$
and $B = (x_1,x_2,\ldots,x_k)$.
\item Assume * and \$ are new symbols.
\item From $(A,B)$ we construct a PCP instance $(C,D)$ with $C = (y_0, y_1,
\ldots, y_{k+1})$ and $D = (z_0,z_1,\ldots,z_{k+1})$ where
\begin{enumerate}
\item $y_i$ is $w_i$ with a * after each symbol in $w_i$, for $i = 1, 2,\ldots,
k$.
\item $z_i$ is $x_i$ with a * before each symbol in $x_i$, for $i = 1, 2,
\ldots, k$.
\item $y_0 = *y_1$ and $y_{k+1} = \$$.
\item $z_0 = z_1$ and $z_{k+1} = *\$$.
\end{enumerate}
\item We can show $i_1, i_2, \ldots, i_m$ is a solution to the given $(A,
B)$-MPCP instance iff $0, i_1, i_2, \ldots, i_m, i_{k+1}$ is a solution to this
constructed $(C,D)$-PCP instance.
\end{itemize}
\end{itemize}

\section{Reducing the Universal Language to MPCP}
\begin{itemize}
\item We can show that given $(M,w)$, an instance of $L_u$, we can reduce this
instance of $L_u$ to an instance $(A, B)$ of the MPCP such that $M$ accepts $w$
iff $(A, B)$ has a solution. We do this by showing that $(A, B)$ simulates the
computation of $M$ on $w$.
\item This shows that both the MPCP and the PCP problems are undecidable.
\end{itemize}

\section{Undecidability of the Ambiguity Problem for CFG's}
\begin{itemize}
\item We can reduce an instance of the PCP problem to an instance of determining
whether a CFG is ambiguous, thereby showing it is undecidable to determine
whether a CFG is ambiguous.
\item We will illustrate the reduction with the following example. Let $(A, B)$
be an instance of the PCP problem with $A = (a, b, ca, abc)$ and $B = (ab, ca,
a, c)$. Let $G$ be the CFG with the productions
\begin{align*}
S &\rightarrow A\,|\,B \\
A &\rightarrow aA1\,|\,bA2\,|\,caA3\,|\,abcA4\,|\,a1\,|\,b2\,|\,ca3\,|\,abc4 \\
B &\rightarrow abB1\,|\,caB2\,|\,aB3\,|\,cB4\,|\,ab1\,|\,ca2\,|\,a3\,|\,c4
\end{align*}
\item There are two distinct leftmost derivations for the string $abcaaabc41321$
because this instance of the PCP problem has a solution.
\end{itemize}
\end{document}
