\documentclass[]{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{indentfirst}
\usepackage{tikz}
\usetikzlibrary{arrows,automata}
\begin{document}

\title{COMS W3261 \\ Computer Science Theory \\ Chapter 1 Notes}
\author{Alexander Roth}
\date{2014-09-06}
\maketitle

\section*{Definitions}
  \begin{description}
    \item[Grammars] A useful model when designing software that processes data 
    with a recursive structure.
    \item[Regular Expressions] Denote the structure of data, especially text
    strings.
  
  \subsection*{Regular Expression Special Characters}
    \begin{description}
      \item[\texttt{\lbrack A-Z\rbrack}] Represents a range of characters from
      capital `A' to capital `Z'.
      \item[\texttt{\lbrack \, \rbrack}] Represents a blank character alone.
      \item[\texttt{*}] Represents ``any number of'' the preceding expression.
      \item[( \, )] Used to group components of the expression. 
    \end{description}
    
  \item[Decidability] What can a computer do at all? Problems that a computer
  can solve are called ``decidable''.
  \item[Intractability] What can a computer do efficiently? The problems that 
  can be solved by a computer using no more time than some slowly growing 
  function of the size of the input are called ``tractable.''
  
  \end{description}
 
\section*{Introduction to Formal Proofs}

  \subsection*{Reduction to Definitions}
    \begin{itemize}
      \item If you are not sure how to start a proof, convert all terms in the
      hypothesis to their definitions.
    \end{itemize}
    \begin{enumerate}
      \item A set $S$ is \emph{finite} if there exists an integer $n$ such that
      $S$ has exactly $n$ elements. We write $\|S\| = n$, where $\|S\|$ is used 
      to denote the number of elements in a set $S$.
      \item If the set $S$ is not finite, we say $S$ is \emph{infinite}.
      Intuitively, an infinite set is a set that contains more than any integer
      number of elements.
      \item If $S$ and $T$ are both subsets of some set $U$, then $T$ is the
      \emph{complement} of $S$ (with respect to $U$) if $S \cup T = U$ and
      $S \cap T = \emptyset$. That is each element of $U$ is in exactly one of
      $S$ and $T$; put another way, $T$ consists of exactly those elements of
      $U$ that are not in $S$.
    \end{enumerate}
    \begin{description}
      \item[Proof by Contradiction] Assume that the conclusion is false. Then
      use that assumption, together with parts of the hypothesis, to prove the
      opposite of one of the given statements of the hypothesis. Shown that it
      is impossible for all parts of the hypothesis to be true and for the
      conclusion to be false at the same time, there is only one possibility.
      The conclusion must be true whenever the hypothesis is true. Thus, the
      theorem is true.
    \end{description}
  
  \subsection*{Ways of Saying ``If-Then''}
    \begin{enumerate}
      \item $H$ implies $C$.
      \item $H$ only if $C$.
      \item $C$ if $H$.
      \item Whenever $H$ holds, $C$ follows.
    \end{enumerate}
      \subsubsection*{Examples}
        \begin{enumerate}
          \item $x \geq 4$ implies $2^x \geq x^2$.
          \item $x \geq 4$ only if $2^x \geq x^2$.
          \item $2^x \geq x^2$ if $x \geq 4$.
          \item Whenever $x \geq 4$, $2^x \geq x^2$ follows.
        \end{enumerate}
    The operator $\rightarrow$ can also take the place of ``if-then''.
    
  \subsection*{If-And-Only-If Statements}
    Forms of the statement ``$A$ if and only if $B$'' include: ``$A$ iff $B$'',
    ``$A$ is equivalent to $B$'' or ``$A$ exactly when $B$''. This statement is
    actually two if-then statements. We prove this by doing:
    \begin{enumerate}
      \item The \emph{if part}: ``if $B$ then $A$,'' and
      \item The \emph{only-if part}: ``if $A$ then $B$,'' which is often stated
      in the equivalent form ``$A$ only if $B$.''
    \end{enumerate}
    The operators $\leftrightarrow$ and $\equiv$ are used to denote 
    ``if-and-only-if'' statements.
  
  \subsection*{How Formal Do Proofs Have to Be?}
    There are certain things that are required in proofs, and omitting them
    surely makes the proof inadequate. For example, any deductive proof that
    uses statements which are not justified by given or previous statements,
    cannot be adequate.

\section*{Additional Forms of Proofs}
  \subsection*{Proving Equivalences About Sets}
    If $E$ and $F$ are two expressions representing sets, the statement
    $E = F$ means that the two sets represented are the same.
    
  \subsection*{The Contrapositive}
    The \emph{contrapositive} of the statement ``if $H$ then $C$'' is ``if not
    $C$ then not $H$''. A statement its contrapositive are either both true or
    both false, so we can prove either to prove the other.
    
  \subsection*{The Converse}
    The \emph{converse} of an if-then statement is the ``other direction''; 
    that is, the converse of ``if $H$ then $C$'' is ``if $C$ then $H$.'' 
    Unlike the contrapositive, which is logically equivalent to the original, 
    the converse is \emph{not} equivalent to the original statement.
  
  \subsection*{Counterexamples}
    Statements that have no parameters, or that apply to only a finite number
    of values of its parameter(s) are called \emph{observations}. It is often
    easier to prove that a statement is not a theorem than to prove it 
    \emph{is} a theorem.
    
\section*{Inductive Proofs}
  In automata theory, inductive proofs are used with recursively defined 
  concepts such as trees and expressions of various sorts (e.g. regular 
  expressions).
  
  \subsection*{Inductions on Integers}
    \begin{itemize}
      \item \emph{The Induction Principle}: If we prove $S(i)$ and we prove 
      that for all $n \geq i$, $S(i)$ implies $S(n + 1)$, then we may conclude
      $S(n)$ for all $n \geq i$.
    \end{itemize}
  
  \subsection*{Structual Inductions}
    When we have a recursive definitions, we can prove theorems about it using
    the following proof form, which is called \emph{structural induction}. Let
    $S(X)$ be a statement about the structures $X$ that are defined by some
    particular recursive definition.
    \begin{enumerate}
      \item As a basis, prove $S(X)$ for the basis structure(s) $X$.
      \item For the inductive step, take a structure $X$ that the recursive
      definition says is formed from $Y_1, Y_2, \ldots, Y_k$. Assume that the
      statements \\ $S(Y_1), S(Y_2),\ldots, S(Y_k)$ hold, and use these to 
      prove $S(X)$.
    \end{enumerate}  
  Our conclusions is that $S(X)$ is true for all $X$.
  
  \subsection*{Mutual Inductions}
    \begin{itemize}
      \item Each of the statements must be proved separately in the basis and
      in the inductive step.
      \item If the statements are ``if-and-only-if,'' then both directions of
      each must be proved, both in the basis and in the induction.
    \end{itemize}
    
\section*{The Central Concepts of Automata Theory}
  \subsection*{Alphabet}
    An \emph{alphabet} is a finite, nonempty set of symbols. Conventionally, we
    use $\Sigma$ for an alphabet.
  
  \subsection*{Strings}
    A \emph{string} is a finite sequence of symbols chosen from some alphabet.
  
    \subsubsection*{The Empty String}
      The \emph{empty string} is the string with zero occurrences of symbols.
      Denoted by $\epsilon$.
 
    \subsubsection*{Length of a String}
      It is often useful to classify strings by their \emph{length}, that is
      the number of positions for symbols in the string. The standard notation
      for the length of a string $w$ is $|w|$.
      
    \subsubsection*{Powers of an Alphabet}
      If $\Sigma$ is an alphabet, we can express the set of all strings of a 
      certain length from that alphabet by using an exponential notation. We
      define $\Sigma^k$ to be the set of strings of length $k$, each of whose
      symbols is in $\Sigma$.
        \begin{itemize}
          \item The set of all strings over an alphabet $\Sigma$ is
          conventionally denoted $\Sigma^*$.
          \item The set of nonempty strings from alphabet $\Sigma$ is denoted
          $\Sigma^+$
          \item $\Sigma^+ = \Sigma^1 \, \cup \, \Sigma^2 \, \cup \, \Sigma^3 \, 
          \cup \, \cdots$.
          \item $\Sigma^* = \Sigma^+ \, \cup \, \{ \epsilon \}$.
        \end{itemize}
      
    \subsubsection*{Concatentation of Strings}
      Let $x$ and $y$ be strings. Then $xy$ denotes the \emph{concatenation} of
      $x$ and $y$, that is, the string formed by making a copy of $x$ and
      following it by a copy of $y$. $\epsilon$ is the \emph{identity for
      concatenation}, since when concatenated with any string it yields the 
      other string as a result.
  
  \subsection*{Languages}
    A set of strings all of which are chosen from some $\Sigma^*$, where 
    $\Sigma$ is a particular alphabet, is called a \emph{language}. If $\Sigma$ 
    is an alphabet, and $L \subseteq \Sigma^*$, then $L$ is a \emph{language 
    over} $\Sigma$. The only important constraint on what can be a language is
    that all alphabets are finite.
    
  \subsection*{Problems}
    A \emph{problem} is the question of deciding whether a given string is a 
    member of some particular language. If $\Sigma$ is an alphabet, and $L$ is 
    a language over $\Sigma$ then the problem $L$ is:
      \begin{itemize}
        \item Given a string $w$ in $\Sigma^*$, decide whether or not $w$ is in   
        $L$.
      \end{itemize}
      
\section*{Summary}
  \begin{description}
    \item[Finite Automata] Finite automata involve states and transitions among 
    states in response to inputs.
    \item[Regular Expressions] These are structural notation for describing the 
    same patterns that can be represented by finite automata.
    \item[Context-Free Grammars] These are an important notation for describing
    the structure of programming languages and related sets of strings.
    \item[Turing Machines] These are automata that model the power of real 
    computers. They allow us to study decidability, the question of what can or 
    cannot be done by a computer. They also let us distinguish tractable 
    problems.
    \item[Deductive Proofs] This basic method of proof proceeds by listing 
    statements that are either given to be true, or that follow logically from 
    some of the previous statements.
    \item[Languages and Problems] A language is a (possibly infinite) set of 
    strings, all of which choose their symbols from one alphabet. When the 
    strings of a language are to be interpreted in some way, the question of 
    whether a string is in the language is sometimes called a problem.
  \end{description}
       
\end{document}