\documentclass[]{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{listings}
\usepackage{multirow}
\usepackage{tikz}
\usepackage{tikz-qtree}
\usepackage{tipa}
\usetikzlibrary{arrows,automata}
\begin{document}
\newcommand*{\xml}[1]{\texttt{<#1>}}
\theoremstyle{definition}
\newtheorem{thm}{Theorem}
\title{COMS W3261 \\ Computer Science Theory \\ Lecture 15 \\ The Universal
Language}
\author{Alexander Roth}
\date{2014 -- 10 -- 27}
\maketitle
\section*{Outline}
\begin{enumerate}
\item Definition of an algorithm
\item The Church-Turing thesis
\item The diagonalization language $L_d$ is not RE
\item Reducing one problem to another
\item The universal language $L_u$ is RE but not recursive
\end{enumerate}

\section{Definition of Algorithm}
\begin{itemize}
\item Surprisingly, there is no universally agreed-upon definition for the term
``algorithm''. Informally, we can think of an algorithm as a collection of well-
defined instructions for carrying out some task.
\item In \emph{The Art of Computer Programming}, Donald Knuth states that an
algorithm should have five properties.
\begin{enumerate}
\item Finiteness: An algorithm must always terminate after a finite number of
steps.
\item Definiteness: Each step of an algorithm must be precisely defined.
\item Input: An algorithm has zero or more inputs.
\item Output: An algorithm has one or more outputs, quantities which have a
specified relation to the inputs.
\item Effectiveness: All of the operations to be performed in an algorithm can
be done exactly and in a finite length of time.
\end{enumerate}
\item In this course we will use a Turing machine that halts on all inputs as
the definition of an algorithm. The term decider is sometimes used for  such a
Turing machine.
\begin{itemize}
\item A language $L$ that can be recognized by an algorithm is said to be
recursive.
\item If a language $L$ is recursive, we say $L$ is decidable.
\item If a language $L$ is not recursive, we say $L$ is undecidable.
\end{itemize}
\item In general, a Turing machine need not halt all inputs. An input on which a
Turing machine never halts is not in the language defined by the Turing machine.
\begin{itemize}
\item A language $L$ that can be recognized by a Turing machine is said to be
recursively enumerable.
\item The term Turing-recognizable language is sometimes used for a recursively
enumerable language.
\item Note that a language may be undecidable because it is not recursive but is
recursively enumerable or because it is not  recursively enumerable.
\end{itemize}
\end{itemize}
\section{The Church-Turing Thesis}
\begin{itemize}
\item A Turing machine can compute a function from an input to an output by
reading the input, making a sequence of moves, and then halting, leaving only
the output of the function on the tape.
\item A recursive function is one that can be computed by a Turing machine that
halts on all inputs.
\item A partial-recursive function is one that can be computed by a Turing
machine that need not halt on all inputs. The output of the function on an input
for which the Turing machine does not halt is said to be undefined.
\item The Church-Turing thesis says that any general way to compute will allow
us to compute only the partial-recursive functions. The Church-Turing thesis is
unprovable because there is no precise definition for ``any general way to
compute.''
\item An informal way of expressing the Church-Turing thesis is that any
function that can be effectively computed can be computed by a Turing machine.
\end{itemize}
\section{The Diagonalization Language $L_d$ is not Recursively Enumerable}
\begin{itemize}
\item We can enumerate all binary strings.
\item We can enumerate all Turing machines.
\item We define $L_d$, the diagonalization language, as follows:
\begin{enumerate}
\item Let $w_1$, $w_2$, $w_3$, $\ldots$ be an enumeration of all binary strings.
\item Let $M_1$, $M_2$, $M_3$, $\ldots$ be an enumeration of all Turing
machines.
\item Let $L_d = \{\,w_i\,|\,w_i$ is not in $L(M_i)\,\}$.
\end{enumerate}
\item Theorem: $L_d$ is not a recursively enumerable language.
\item Proof:
\begin{itemize}
\item Suppose $L_d = L(M_i)$ for some TM $M_i$.
\item This gives rise to a contradiction. Consider what $M_i$ will do on the
input $w_i$.
\begin{itemize}
\item If $M_i$ accepts $w_i$, then by definition $w_i$ cannot be in $L_d$.
\item If $M_i$ does not accept $w_i$, then by definition $w_i$ is in$L_d$.
\end{itemize}
\item Since $w_i$ can neither be in $L_d$ nor not be in $L_d$, we must conclude
there is no Turing machine that can define $L_d$.
\end{itemize}
\end{itemize}
\section{Reducing One Problem to Another}
\begin{itemize}
\item If we have an algorithm to convert instance of a problem $P_1$ to
instances of a problem $P_2$ that have the same answer, then we say that $P_1$
reduces to $P_2$.
\item A reduction from $P_1$ to $P_2$ must turn every instance of $P_1$ with a
yes answer to an instance of $P_2$ with a yes answer, and every instance of
$P_1$ with a no answer to an instance of $P_2$ with a no answer.
\item We will frequently use this technique to show that problem $P_2$ is ashard
as problem $P_1$.
\item The direction of the reduction is important.
\item For example, if there is a reduction from $P_1$ to $P_2$ and if $P_1$ is
not recursive, then $P_2$ cannot be recursive.
\item Similarly, if there is a reduction from $P_1$ to $P_2$ and if $P_1$ is not
recursively enumerable, then $P_2$ cannot be recursively enumerable.
\end{itemize}
\section{The Universal Language $L_u$ is RE but not Recursive}
\begin{itemize}
\item $L_u$, the universal language, is the set of binary strings that encode a
pair $(M, w)$ consisting of a Turing machine and an input string accepted by
that Turing machine. That is,
\begin{itemize}
\item $L_u = \{\,(M,w)\,|\,M$ is an encoding of a Turing machine, $w$ is an
encoding of a binary string, and $w$ is in $L(M)\,\}$.
\end{itemize}
\item Theorem: $L_u$ is recursively enumerable.
\item Proof:
\begin{itemize}
\item We can construct a Turing machine $U$, called the universal Turing
machine, to recognize $L_u$.
\item It is easiest to think of $U$ as a multi-tape TM.
\begin{enumerate}
\item One tape holds the input with the encodings of $M$ and $w$.
\item A second tape is used to simulate $M$'s input tape.
\item A third tape is used to keep track of $M$'s state.
\item A fourth tape is used as a scratch tape.
\end{enumerate}
\item To simulate a move of $M$, $U$ searches for a transition on the current
state of $M$ (stored  on tape 3) and the current state tape symbol of $M$
(stored at the position on tape 2 scanned by $U$).
\begin{enumerate}
\item $U$ changes the contents of tape 3 to record the new state.
\item $U$ changes the tape symbol under $M$'s simulated tape head on tape 2.
\item $U$ moves $M$'s tape head left or right on tape 2.
\end{enumerate}
\item If $M$ enters its final state, $U$ accepts the original input $(M, w)$.
\item $L_u = L(U)$.
\end{itemize}
\item Theorem: $L_u$ is not recursive.
\item Proof:
\begin{itemize}
\item Suppose $L_u$ were recursive. Then there exists a TM $M$ that accepts the
complement of $L_u$.
\item Then we can transform $M$ into a TM $M^\prime$ that accepts $L_d$ as
follows:
\begin{itemize}
\item $M^\prime$ transforms its input string $w$ into a pair $(w,w)$.
\item $M^\prime$ simulates $M$ on $(w,w)$ assuming the first $w$ is an encoding
of a TM $M_i$ and the second $w$ is an encoding of a binary string $w_i$. Since
$M$ accepts the complement of $L_u$, it will accept $(w,w)$ iff $M_i$ does not
accept $w_i$.
\end{itemize}
\item Thus, $M^\prime$ accepts $w$ iff $w$ is in $L_d$. But we have previously
shown there does not exist a TM that defines $L_d$.
\item We conclude $L_u$ is not recursive.
\end{itemize}
\end{itemize}
\end{document}
