\documentclass[]{article}
\usepackage[all]{xy}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{enumitem}
\usepackage{indentfirst}
\usepackage{hhline}
\usepackage{listings}
\usepackage{multirow}
\usepackage{tikz}
\usepackage{tikz-qtree}
\usepackage{tipa}
\usetikzlibrary{arrows,automata}
\begin{document}

\newtheorem{thm}{Theorem}
\title{COMS W3261 \\ Computer Science Theory \\ Homework \#4}
\author{Alexander Roth}
\date{2014 -- 11 -- 12}
\maketitle

\section*{Problems}
\begin{enumerate}
\item Classify each of the following languages as being (i) recursive, (ii)
recursively enumerable but not recursive, or (iii) not recursively enumerable.
In each case briefly justify your answer.
\begin{enumerate}
\item $\{\,D\,|\,D$ is a deterministic finite automaton that accepts at least
one string with an equal number of $a$'s and $b$'s$\,\}$.
\item[\emph{Solution}:] This language is recursive. We can simulate a DFA $D$
that accepts at least one string with an equal number of $a$'s and $b$'s, and we
can create a Turing machine $M$ that simulates this DFA. Thus, $M$ will always
halt whenever $D$ halts, which occurs either when it accepts the string or
rejects said string.

\item $\{\,(D_1,D_2)\,|\,D_1$ and $D_2$ are deterministic finite automata and
$L(D_1) = L(D_2)\,\}$
\item[\emph{Solution}:] This language is recursive. We can construct a Turing
machine $M$ that simulates the table-filling algorithm that determines if two
DFA's are equivalent. Thus, we can check for equivalence between $D_1$ and
$D_2$; if we find that $L(D_1) = L(D_2)$, $M$ will halt and accept, otherwise it
will halt without accepting.

\item $\{\,(M,w)\,|\,M$ is a Turing machine that halts on input $w\,\}$.
\item[\emph{Solution}:] This language is recursively enumerable. We can
construct a Turing machine $Q$ that simulates the Turing machine $M$. Thus, we
can simulate $M$ on $Q$. However, we do not know if $M$ will ever halt; thus,
$Q$ may never halt unless $M$ halts and accepts. Therefore, it is recursively
enumerable.

\item $\{\,(M_1,M_2)\,|\,M_1$ and $M_2$ are Turing machines and $L(M_1) =
L(M_2)\,\}$
\item[\emph{Solution}:] This language is not recursively-enumerable. To test the
equivalency between $M_1$ and $M_2$, we must check that $M_1$ and $M_2$ accept
and reject the same input strings. Typically, the scenario to test equivalence
between $M_1$ and $M_2$ would be recursively enumerable if there would most
likely be a finite set of accepting strings. However, suppose that one of the
machines, $M_1$ in this case, has an infinite set of strings to which it halts
and accepts. Thus, we cannot even say for certain if the machines are equivalent
since we will continue to check halting strings in an infinite set. Thus, the
language is not recursively-enumerable.
\end{enumerate}

\item Boolean expressions
\begin{enumerate}
\item Describe an algorithm in high-level pseudocode to evaluate a boolean
expression given a truth assignment to its variables. Describe how your
algorithm evaluates the boolean expression $x \vee \neg(y \wedge z)$ for the
truth assignment $T(x) = 0$, $T(y) = 0$, $T(z) = 1$. (1 is true, 0 is false)
\item[\emph{Solution}:] We can construct a parse tree where the leafs of the
tree are all literals of the boolean expression, and any negation we have found
on the way gets pushed down to the literals. From there, we can traverse up the
tree, evaluating the truth values as we go. Once we reach the top of the tree,
we have figured out the truth values for the whole boolean expression.

We can construct a parse tree for $x \vee \neg(y \wedge z)$ as follows: \\
\Tree [.$\vee$ [.$x$ ] [.$\vee$ [.$\overline{y}$ ] [.$\overline{z}$ ] ] ] \\
Now, we read up the tree. Thus, $(\overline{y} \vee \overline{z})$ becomes $(1
\vee 0)$, which evaluates to $1$. From there, we have $(x \vee 1)$, which is
true.

\item What is the time complexity of your algorithm?
\item[\emph{Solution}:] The time complexity for this algorithm is $O(n)$. We
must first construct a parse tree, which takes $O(n)$, and the read up the parse
tree, which is again $O(n)$ time. Thus, $O(n + n)$ is $O(n)$ overall.

\item Describe an algorithm in high-level pseudocode to determine whether a
boolean expression always evaluates to true on all truth assignments to its
variables (is a tautology). Illustrate how your algorithm performs on the
boolean expression $x \vee \neg (y \wedge x)$.
\item[\emph{Solution}:] In order to determine if a boolean expression always
evaluates to true, we shall test with every possible truth assignment for every
variable using the algorithm described in 3a. If in the course of this
algorithm, we find that one of the combinations does not evaluate to true
overall, we know that it cannot be a tautology.

For example, $x \vee \neg(y \wedge x)$ has 4 possible solutions that prove it is
a tautology \\
\begin{tabular}{|c|c|c|c|}
\hline
$x$ & $y$ & $\neg(y\wedge x)$ & $x \vee \neg(y \wedge x)$ \\ \hline
0   & 0   & 1                 & 1                         \\ \hline
0   & 1   & 1                 & 1                         \\ \hline
1   & 0   & 1                 & 1                         \\ \hline
1   & 1   & 0                 & 1                         \\ \hline
\end{tabular}

\item What is the time complexity of your algorithm?
\item[\emph{Solution}:] The time complexity to build the parse tree is $O(2^n)$
as
it is necessary to test each truth value for $n$ variables. Therefore, there are
two possible choices for each variable.

\end{enumerate}

\item The game PEBBLES is played on an $k \times n$ chessboard. Initially each
square of the chessboard has a black pebble, or a white pebble, or no pebble.
You play the game by removing pebbles one at a time. You win the game if you can
end up with a board in which each column contains only pebbles of a single color
and each row contains at least one pebble.
\begin{enumerate}
\item Show that the set of winnable PEBBLES game is in NP. Hint: Describe a
nondeterministic polynomial-time algorithm to  determine whether a given PEBBLES
board is winnable.
\item[\emph{Solution}:] The set of winnable PEBBLES boards is in $\mathcal{NP}$.
Construct a nondeterministic polynomial algorithm that will determine if
a color should be completely removed from a given row. After each of the rows
has been guessed, the algorithm determines if the board is winnable. This
algorithm removes every possible subset of pebbles and accepts the only winnable
one.

\item Given a boolean expression $E$ in 3-CNF with $k$ clauses and $n$
variables, construct the following $k \times n$ board: If literal $x_i$ is in
clause $c_j$, put a black pebble in column $x_i$, row $c_j$. If literal $\neg
x_i$ is in clause $c_j$, put a white pebble in column $x_i$, row $c_j$. Show
that $E$ is satisfiable if and only if this PEBBLES game is winnable.
\item[\emph{Solution}:] We must create a mapping between the PEBBLES game and
the 3-CNF $E$ and show that the map is bidirectional. If we were to map the
PEBBLE game to a CNF, each row would be a clause while each column would be a
literal. A literal cannot have more than one truth value, which corresponds to
either having one column of a single color (black or white). From the 3-CNF $E$,
we see that a black pebble will represent true and the white pebble will
represent false. In the game, we will remove all white pebbles when the literal
is true and remove all black pebble when the literal is false.

If a row in the game is empty, this is equivalent to the saying the clause is a
fallacy. If a column is empty, then this literal does not appear in boolean
expression; however, this cannot be the case as there must be a truth assignment
for all $n$ variables, as per the instructions.

Therefore, we see that the PEBBLES game is equivalent to the 3-CNF $E$ through a
mapping of its characteristics and requirements for satisfiability.

\item What can you conclude from showing (a) and (b)?
\item[\emph{Solution}:] Any 3-CNF problem is $\mathcal{NP}$-complete; thus, $E$
is NP-complete. From section (a), we know that the game is in $\mathcal{NP}$.
Since we were able to reduce the PEBBLES game to the 3-CNF satisfiability
problem, we are able to show that the PEBBLES game is also
$\mathcal{NP}$-complete.
\end{enumerate}

\item A polynomial-time verifier for a language $L$ is an algorithm $V$ such
that $L = \{\,w\,|\,V$ accepts $(w,c)$ in polynomial time for some
polynomial-length string $c$ called a certificate or proof of membership in
$L\,\}$. Show that NP is equivalent to the class of languages that have
polynomial-time verifiers.
\item[\emph{Solution}:] We must convert the polynomial time verifier into a
polynomial time Nondeterministic Turing machine, and we must convert the
Nondeterministic Turing machine into a polynomial time verifier. The Turing
machine simulates the verifier by guessing the certificate. Likewise, we shall
use the accepting branch of the machine as the accepting branch for the
verifier.

Let us assume that $L$ is in $\mathcal{NP}$. We have a verifier $V$ that
verifies $L$ in polynomial time. We can construct a Nondeterministic Turing
machine $M$ such that given input $w$, we nondeterministically select a
certificate $c$ which is at most polynomial in length. Thus, we test $V$ on the
input of $(w, c)$. If $V$ accepts, $M$ should accept; otherwise, $M$ will
reject.

Assume that $L$ is decided by a polynomial time Nondeterministic Turing machine
$M$. We can construct a verifier $V$ such that, given input $(w, c)$, $V$ checks
whether $c$ becomes an accepting branch of $M$ given input $w$. If $c$ is an
accepting branch, $M$ accepts; otherwise, $M$ rejects $w$.

\item The classes P and NP
\begin{enumerate}
\item Show that if language $A$ is $\mathcal{NP}$-complete and $A$ is in
$\mathcal{P}$, then $\mathcal{P=NP}$.
\item[\emph{Solution}:] Suppose $A$ is both $\mathcal{NP}$-complete and in
$\mathcal{P}$. Then all languages $L$ is $\mathcal{NP}$ reduces in polynomial-
time to $A$. If $A$ is in $\mathcal{P}$, then $L$ is in $\mathcal{P}$.

\item Show that if $A$ is $\mathcal{NP}$-complete and $A$ is polynomially
reducible to a language $B$ in $\mathcal{NP}$, then $B$ is
$\mathcal{NP}$-complete.
\item[\emph{Solution}:] We need to show that every language $L$ in
$\mathcal{NP}$ polynomial-time reduces to $B$. We know that there is a
polynomial-time reduction of $L$ to $A$, since $A$ is already NP-complete. Thus,
a string $w$ in $L$ of length $N$ is converted to a string $x$ in $A$ of length
at most $p(n)$ where $p(n)$ is the polynomial time reduction.

We know that there is a polynomial-time reduction of $A$ to $B$; this reduction
takes polynomial time $q(m)$. This reduction transforms $x$ to some string $y$
in $B$, taking time at most $q(p(n))$. Thus, the transformation of $w$ to $y$
takes time at most $p(n) + q(p(n))$, which is a polynomial. Therefore, $L$ is
polynomial-time reducible to $B$. Since $L$ could be any language in
$\mathcal{NP}$, we know that all of $\mathcal{NP}$ polynomial-time reduces to
$B$; thus, $B$ is NP-complete.

\end{enumerate}
\end{enumerate}

\end{document}
